Fractional Decomposition Tree Algorithm: A tool for studying the integrality gap of Integer Programs

### 2022-05-05 FDT2020/Introduction.tex
Comments from Bob

For example consider the minimum cost \textsc{2-edge-connected multi-subgraph problem (2EC)}: Given a graph $G=(V,E)$ and $c\in \mathbb{R}^E_{\geq 0}$, 2EC asks for the minimum cost 2-edge-connected subgraph of $G$, with multi-edges allowed. A graph is 2-edge-connected if there are at least two edge-disjoint paths between every pair of vertices.  An IP formulation for this problem, known as the {\em subtour-elimination}, is
 \begin{equation}\min \{cx: \sum_{e\in \delta(U)}x_e \geq 2 \mbox{ for } \emptyset \subsetneq U \subsetneq V,\; x\in \{0,1,2\}^{E}\}, 
 \label{eq:subtour}
 \end{equation}
where $\delta(U)$ for vertex subset $U$ is the set of edges that cross the cut defined by $U$.  That is, each $e \in \delta(U)$ has one endpoint in $U$ and the other endpoint in $V-U$. We denote the linear relaxation of this IP as $\subtour(G)$. Notice that in this case, $G$ defines the polyhedron, while the pair of $G$ and $c$ defines the instance of 2EC. Let $I_G$ be the polyhedron instance defined by $\subtour(G)$. Then $g(I_G)$ is the worst-case integrality gap 2EC on $G$ over all non-negative cost functions and integrality gap of 2EC is $\sup_{G} g(I_G)$.

@@

Formally, for some positive integer $k$, given a set of points $S\subseteq \mathbb{R}^k$ and a vector $x\in \mathbb{R}^k$, we say \textit{$x$ \underline{is} as a convex combination of points in $S$} if there exists a positive integer $\ell$ and $\theta \in [0,1]^\ell$, where $\sum_{i=1}^{\ell}\theta_i =1$ and $\tilde{x}^i\in S$ for $i\in [k]$ such that $\sum_{i=1}^{k}\theta_i \tilde{x}^i\leq x$. If such a convex combination can be found in time polynomial in the encoding size of the definition of $S$ and $x$, then we say  \textit{$x$ \underline{can be written} as a convex combination of points in $S$}

@@

\begin{thm}[Carr, Vempala \cite{Carr2004}] \label{CV2}
	We have $g(I) \leq C$ if and only if for $x\in P(I)$ the vector $Cx$ can be expressed as a convex combination of points in $\dom(S(I))$. 
\end{thm}
Goemans \cite{goemansblocking} first introduced Theorem \ref{CV2} for blocking-type polyhedra. While there is an exact algorithm for problems with gap $1$, as stated in Proposition~\ref{cara}, Theorem~\ref{CV2} is existential, with no construction. Proposition~\label{cara} would not work if we drop the non-negativity constraints on the cost function $c$ in the definition of integrality gap (Definition~\ref{IGofPoly}).
%In contrast to Proposition \ref{cara} which implies exact algorithms for problems with a gap of 1, Theorem \ref{CV} does not imply an approximation algorithm, since it does not suggest how to find such a convex combination in polynomial time.
%This points to an interesting open question. 
% We show later, for $I$ with $g(I)<\infty$, the notation of dominant is in fact useful.
\iffalse

@@

In Section~\ref{sec:binaryfdt} we introduce the \textit{Fractional Decomposition Tree} (FDT) which is polynomial time algorithm. Given a point $x\in P(I)$ FDT produces a convex combination of feasible points in $S(I)$ that are dominated coordinatewise by a ``factor" $C$ times $x$.
% in the coordinates corresponding to $x$.
If $C = g(I)$, FDT would have been optimal. However we can only guarantee $C\leq g(I)^{|\spp(x)|}$. FDT iteratively solves linear programs that are about the same size as the description of $P(I)$.

\begin{restatable}{thm}{binaryFDT}
	\label{binaryFDT}
	Assume $g(I) 	<\infty$. For $x^*\in P(I)$, using 	the Fractional Decomposition Tree (FDT) algorithm we can write $x$ as a convex combination of points in $S(I)$, where $C\leq g(I)^{|\spp(x^*)|}$.
\end{restatable}

@@

A keen reader might ask why Theorem \ref{domtoIP} does not imply that P = NP, as one iteratively find a feasible solution $x^*\in S(I)$, and add the constraint $x\leq cx^*-\epsilon$ for some small enough $\epsilon>0$ to $I$ until DomToIP fails to find a feasible solution. However, note that adding constraints to $I$ might result in $g(I)$ becoming infinite, in which case DomToIP would fail to find a feasible solution to the IP.

@@

Here is how the FDT algorithm works at a high level for an instance $I$ with LP feasible region $P(I)$: in iteration $i-1$ the algorithm computes a convex combination of vectors in $\mathcal{D}(P(I))$ that have a 0 or 1 value for coordinates indexed $1,\ldots,i-1$. Let $y$ be a vector in this convex combination from iteration $i-1$. In iteration $i$, we solve a linear-programming problem that gives us $\gamma_0,\gamma_1\in [0,1]$ and $y^0,y^1\in \mathcal{D}(P(I))$ such that $y\geq \gamma_0 y^0 + \gamma_1 y^1$, $\gamma_0+\\gamma_1\leq g(I)$, $y^0_i=0$ and $y^1_i=1$. We then replace $y$ in the convex combination with $\frac{\gamma_0}{\gamma_0+\gamma_1}y^0 +\frac{\gamma_1}{\gamma_0+\gamma_1}y^1$.  Repeating this for every vector in the convex combination from the previous iteration yields a new convex combination of points in $\dom(P(I))$. It is ``more'' integral because now all vectors in the convex combination are integral in their first $i$ coordinates. If in any iteration there are too many points in the convex combination,  we solve a linear-programming problem that ``prunes'' the convex combination. At the end we have a convex combination of integer solutions $\mathcal{D}(P(I))$. For each such solution $z$ we invoke the DomToIP algorithm
% CAP: forward pointer earlier now
% (see Section~\ref{sec:domTOIP})
to find $z'\in S(I)$ where $z'\leq z$.

One can extend the FDT algorithm for binary IPs into covering\footnote{A covering IP has nonnegative constraint matrix, objective coefficients and right-hand side ($A,b,c$).} $\{0,1,2\}$ IPs by losing a factor $2$ on top of the loss for FDT. To eradicate this extra factor, we must treat the coordinate $i$ with $x_i=1$ differently. In Section~\ref{sec:2EC} we focus on the 2-edge-connected multi-subgraph graph problem (2EC) defined above.  Its subtour-elimination LP relaxation is given in (\ref{eq:subtour}).
%}: Given a graph $G=(V,E)$ and $c\in \mathbb{R}^{E}_{\geq 0}$ find a 2-edge-connected multi-subgraph of $G$ with minimum cost. The natural linear programming relaxation for this problem is 
%\begin{equation}
%\min \{cx \; : \; x(\delta(U))\geq 2 \; \text{ for }\emptyset \subset U \subset V, \; x\in [0,2]^E\}
%\end{equation}
For input graph $G$, let $\subtour(G)$ denote the feasible region of this LP. Let $\2ec(G)$ be the convex hull of incidence vectors\footnote{The incidence vector of a 2-edge-connected multi-subgraph $F$, denoted by $\chi^F$ has an element for each edge $e$, with $x_e \in \{0,1,2\}$ indicating the number of times edge $e$ appears in the solution.} of 2-edge-connected multi-subgraphs of graph $G$. Following the definition in (\ref{gapproblem}) have
\begin{equation}
g(\2ec) = \sup_{c\geq 0 , G}\frac{\min_{x\in \2ec(G)} cx}{\min_{x\in \subtour(G)} cx}.
\end{equation}

@@

%\arash{I added this figure \ref{fig:CVpoint} but maybe we can put it later in the experiments section to make the introduction shorter?} \cindy{Yes that would be fine, if the reader could also use that figure then.  They could page forward to see the figure when reading the introduction.}
%\arash{I changed the paragraph below a little bit, can you check?} \cindy{Looks good. I just changed one sentence to the active voice.  If you're OK with that, you can delete these 2 comments. Thanks!}
Carr and Vempala \cite{Carr2004} showed that $g(\2ec)$ is achieved for instances where the optimal solution to $\min_{x\in \subtour(G)}cx$ is a Carr-Vempala point. Multiple groups of researchers have conjectured that $g(\2ec)\leq \frac{6}{5}$ (see \cite{alexander2006integrality,boydlegault,hn19}, but the only known upper bound on $g(\2ec)$ stands at $\frac{3}{2}$ \cite{wolsey}. We show that the integrality gap is at most $\frac{6}{5}$ for Carr-Vempala points with at most 12 vertices on the cycle formed by the fractional edges, thereby confirming the conjecture for these instances. Note that the number of vertices in these instances can be arbitrarily high since the paths of edges with $x$-value 1 can be arbitrarily long.
\cindy{This sounds like a contribution of the paper.} \arash{Yes, particlarly since a similar research question have been studied by \cite{alexander2006integrality} and \cite{TSPcompute}}
\cindy{We probably need an explicit list of contributions.  It could be another subsection of the introduction.  This will help the reviewer and eventually, if published, the reader.}
\subsection{Contribution summary}
In summary, this papers's contributions are:
\begin{itemize}
\item We give an algorithm to express any feasible point for the LP relaxation of a binary IP
%(or a \{0,1,2\} IP) 
as a convex combination of feasible solutions, provided the IP integrality gap is bounded.  This is the Fractional Decomposition Tree (FDT) algorithm.
\item We give a simple  algorithm to give a feasible solution for any binary IP with bounded integrality gap. If this algorithm fails, then the integrality gap is infinite.
\item We experimentally show that FDT can give good approximate solutions to small instances of vertex-cover and tree augmentation problems.
\item We show that the integrality gap for 2EC (minimum 2-edge-connected sub-multigraph) is at most $\frac{6}{5}$ for Carr-Vempala points with at most $12$ vertices on the cycle formed by the fractional edges, matching the lower bound for integrality gap for 2EC for these instances.
\item We demonstrate (with 2EC) how FDT exeriments can support theoretical work on integrality gaps for specific problems. In particular, we computationally prove an upper bound on small instances of 2EC with a specific structure.
\item We demonstrate (with VC) how FDT can be used to find feasible solutions of improved quality compared to those from the feasiblity pump \cite{fp1}.
\end{itemize}

>>>>

For example consider the minimum cost \textsc{2-edge-connected multi-subgraph problem (2EC)}: Given a graph $G=(V,E)$ and $c\in \mathbb{R}^E_{\geq 0}$, 2EC asks for the minimum cost 2-edge-connected subgraph of $G$, with multi-edges allowed. A graph is 2-edge-connected if there are at least two edge-disjoint paths between every pair of vertices.  An IP formulation for this problem, known as the {\em subtour-elimination}, is
 \begin{equation}\min \{cx: \sum_{e\in \delta(U)}x_e \geq 2 \mbox{ for } \emptyset \subsetneq U \subsetneq V,\; x\in \{0,1,2\}^{E}\}, 
 \label{eq:subtour}
 \end{equation}
where $\delta(U)$ for vertex subset $U$ is the set of edges that cross the cut defined by $U$.  That is, each $e \in \delta(U)$ has one endpoint in $U$ and the other endpoint in $V-U$. We denote the linear relaxation of this IP as $\subtour(G)$. Notice that in this case, $G$ defines the polyhedron, while the pair of $G$ and $c$ defines the instance of 2EC. Let $I_G$ be the polyhedron instance defined by $\subtour(G)$. Then $g(I_G)$ is the worst-case integrality gap of 2EC on $G$ over all non-negative cost functions and the integrality gap of 2EC is $\sup_{G} g(I_G)$.

@@

Formally, for some positive integer $k$, given a set of points $S\subseteq \mathbb{R}^k$ and a vector $x\in \mathbb{R}^k$, we say \textit{$x$ \underline{is/dominates} a convex combination of points in $S$} if there exists a positive integer $\ell$ and $\theta \in [0,1]^\ell$, where $\sum_{i=1}^{\ell}\theta_i =1$ and $\tilde{x}^i\in S$ for $i\in [\ell]$ such that $\sum_{i=1}^{\ell}\theta_i \tilde{x}^i\leq x$. If such a convex combination can be found in time polynomial in the encoding size of the definition of $S$ and $x$, then we say  \textit{$x$ \underline{can be written} as a convex combination of points in $S$}

@@

\begin{thm}[Carr, Vempala \cite{Carr2004}] \label{CV2}
	We have $g(I) \leq C$ if and only if for $x\in P(I)$ the vector $Cx$ dominates a convex combination of points in $\dom(S(I))$. 
\end{thm}
Goemans \cite{goemansblocking} first introduced Theorem \ref{CV2} for blocking-type polyhedra. While there is an exact algorithm for problems with gap $1$, as stated in Theorem~\ref{CV2} is existential, with no construction. Proposition~\label{cara} would not work if we drop the non-negativity constraints on the cost function $c$ in the definition of integrality gap (Definition~\ref{IGofPoly}).
%In contrast to Proposition \ref{cara} which implies exact algorithms for problems with a gap of 1, Theorem \ref{CV} does not imply an approximation algorithm, since it does not suggest how to find such a convex combination in polynomial time.
%This points to an interesting open question. 
% We show later, for $I$ with $g(I)<\infty$, the notation of dominant is in fact useful.
\iffalse

@@

In Section~\ref{sec:binaryfdt} we introduce the \textit{Fractional Decomposition Tree} (FDT) which is a polynomial time algorithm. Given a point $x\in P(I)$ FDT produces a convex combination of feasible points in $S(I)$ that are dominated coordinatewise by a ``factor" $C$ times $x$.
% in the coordinates corresponding to $x$.
If $C = g(I)$, FDT would have been optimal. However we can only guarantee $C\leq g(I)^{|\spp(x)|}$. FDT iteratively solves linear programs that are about the same size as the description of $P(I)$.

\begin{restatable}{thm}{binaryFDT}
	\label{binaryFDT}
	Assume $g(I) 	<\infty$. For $x^*\in P(I)$, using 	the Fractional Decomposition Tree (FDT) algorithm we can write $Cx^*$ as a convex combination of points in $S(I)$, where $C\leq g(I)^{|\spp(x^*)|}$.
\end{restatable}

@@

A keen reader might ask why Theorem \ref{domtoIP} does not imply that P = NP, as one iteratively find a feasible solution $x^*\in S(I)$, and add the constraint $cx\leq cx^*-\epsilon$ for some small enough $\epsilon>0$ to $I$ until DomToIP fails to find a feasible solution. However, note that adding constraints to $I$ might result in $g(I)$ becoming infinite, in which case DomToIP would fail to find a feasible solution to the IP.

@@

Here is how the FDT algorithm works at a high level for an instance $I$ with LP feasible region $P(I)$: in iteration $i-1$ the algorithm computes a convex combination of vectors in $\mathcal{D}(P(I))$ that have a 0 or 1 value for coordinates indexed $1,\ldots,i-1$. Let $y$ be a vector in this convex combination from iteration $i-1$. In iteration $i$, we solve a linear-programming problem that gives us $\gamma_0,\gamma_1\in [0,1]$ and $y^0,y^1\in \mathcal{D}(P(I))$ such that $y\geq \gamma_0 y^0 + \gamma_1 y^1$, $\gamma_0+\gamma_1\leq g(I)$, $y^0_i=0$ and $y^1_i=1$. We then replace $y$ in the convex combination with $\frac{\gamma_0}{\gamma_0+\gamma_1}y^0 +\frac{\gamma_1}{\gamma_0+\gamma_1}y^1$.  Repeating this for every vector in the convex combination from the previous iteration yields a new convex combination of points in $\dom(P(I))$. It is ``more'' integral because now all vectors in the convex combination are integral in their first $i$ coordinates. If in any iteration there are too many points in the convex combination,  we solve a linear-programming problem that ``prunes'' the convex combination. At the end we have a convex combination of integer solutions $\mathcal{D}(P(I))$. For each such solution $z$ we invoke the DomToIP algorithm
% CAP: forward pointer earlier now
% (see Section~\ref{sec:domTOIP})
to find $z'\in S(I)$ where $z'\leq z$.

One can extend the FDT algorithm for binary IPs into covering\footnote{A covering IP has nonnegative constraint matrix, right-hand side and objective coefficients ($A,b,c$).} $\{0,1,2\}$ IPs by losing a factor $2$ on top of the loss for FDT. To eradicate this extra factor, we must treat the coordinate $i$ with $x_i=1$ differently. In Section~\ref{sec:2EC} we focus on the 2-edge-connected multi-subgraph graph problem (2EC) defined above.  Its subtour-elimination LP relaxation is given in (\ref{eq:subtour}).
%}: Given a graph $G=(V,E)$ and $c\in \mathbb{R}^{E}_{\geq 0}$ find a 2-edge-connected multi-subgraph of $G$ with minimum cost. The natural linear programming relaxation for this problem is 
%\begin{equation}
%\min \{cx \; : \; x(\delta(U))\geq 2 \; \text{ for }\emptyset \subset U \subset V, \; x\in [0,2]^E\}
%\end{equation}
For input graph $G$, let $\subtour(G)$ denote the feasible region of this LP. Let $\2ec(G)$ be the convex hull of incidence vectors\footnote{The incidence vector of a 2-edge-connected multi-subgraph $F$, denoted by $\chi^F$ has an element for each edge $e$, with $x_e \in \{0,1,2\}$ indicating the number of times edge $e$ appears in the solution.} of 2-edge-connected multi-subgraphs of graph $G$. Following the definition in (\ref{gapproblem}) have
\begin{equation}
g(\2ec) = \sup_{c\geq 0 , G}\frac{\min_{x\in \2ec(G)} cx}{\min_{x\in \subtour(G)} cx}.
\end{equation}

@@

%\arash{I added this figure \ref{fig:CVpoint} but maybe we can put it later in the experiments section to make the introduction shorter?} \cindy{Yes that would be fine, if the reader could also use that figure then.  They could page forward to see the figure when reading the introduction.}
%\arash{I changed the paragraph below a little bit, can you check?} \cindy{Looks good. I just changed one sentence to the active voice.  If you're OK with that, you can delete these 2 comments. Thanks!}
Carr and Vempala \cite{Carr2004} showed that $g(\2ec)$ is achieved for instances where the optimal solution to $\min_{x\in \subtour(G)}cx$ is a Carr-Vempala point. Multiple groups of researchers have conjectured that $g(\2ec)\leq \frac{6}{5}$ (see \cite{alexander2006integrality,boydlegault,hn19}, but the only known upper bound on $g(\2ec)$ stands at $\frac{3}{2}$ \cite{wolsey}. We show that the integrality gap is at most $\frac{6}{5}$ for Carr-Vempala points with at most 12 vertices on the cycle formed by the fractional edges, thereby confirming the conjecture for these instances. Note that the number of vertices in these instances can be arbitrarily high since the paths of edges with $x$-value 1 can be arbitrarily long.
\cindy{This sounds like a contribution of the paper.} \arash{Yes, particlarly since a similar research question have been studied by \cite{alexander2006integrality} and \cite{TSPcompute}}
\cindy{We probably need an explicit list of contributions.  It could be another subsection of the introduction.  This will help the reviewer and eventually, if published, the reader.}
\subsection{Contribution summary}
In summary, this papers's contributions are:
\begin{itemize}
\item We give an algorithm to express any feasible point for the LP relaxation of a binary IP
%(or a \{0,1,2\} IP) 
as a convex combination of feasible solutions, provided the IP integrality gap is bounded, when scaled by a factor depending on the integrality gap of the IP formulation.  This is the Fractional Decomposition Tree (FDT) algorithm.
\item We give a simple  algorithm to give a feasible solution for any binary IP with bounded integrality gap. If this algorithm fails, then the integrality gap is infinite.
\item We experimentally show that FDT can give good approximate solutions to small instances of vertex-cover and tree augmentation problems.
\item We show that the integrality gap for 2EC (minimum 2-edge-connected sub-multigraph) is at most $\frac{6}{5}$ for Carr-Vempala points with at most $12$ vertices on the cycle formed by the fractional edges, matching the lower bound for integrality gap for 2EC for these instances.
\item We demonstrate (with 2EC) how FDT experiments can support theoretical work on integrality gaps for specific problems. In particular, we computationally prove an upper bound on small instances of 2EC with a specific structure.
\item We demonstrate (with VC) how FDT can be used to find feasible solutions of improved quality compared to those from the feasiblity pump \cite{fp1}.
\end{itemize}

### 2022-05-05 FDT2020/main-section.tex
Comments from Bob

Consider an IP instance $I=(A,b)$. Define sets $S(I)$ and $P(I)$ as in (\ref{S}) and (\ref{P}), respectively. Assume $S(I)\subseteq \{0,1\}^n$ and $P(I)\subseteq [0,1]^n$. For simplicity in the notation we assume an instance $I$ and denote $P(I),S(I),$ and $g(I)$ by $P$, $S$, and $g$ for this section and the next section. Also, for both sections let $x^*$ be the optimal solution to the LP formulation and assume $t=|\spp(x^*)|$. Reorder variables as necessary to put all the nonzero variables into the first $t$ indices. So we can assume $x^*_i = 0$ for $i=t+1,\ldots,n$.

Before diving into the main results, we state the following simple observation that we will use several times in the proofs of this section and the next section.
\begin{observation}\label{dom01}
Any point $x\in [0,1]^n$ that is a convex combination of points in $\dom(S)$, is also a convex combination of points in $\dom(S) \cap [0,1]^n$
\end{observation}
\begin{proof}
This is because if there exists ${z}^i\in \dom(S)$ and $\theta_i\geq 0$ for $i\in [k]$ such that $\sum_{i=1}^{k} \theta_i = 1$ and $\sum_{i=1}^{k}\theta_i {z}^i \leq gz$, then 
we let ${y}^i$ be the coordiante-wise minimum of vector ${z}^i$ and $\{1\}^n$ for $i\in [k]$ and we have ${y}^i\in \dom(S)$ and $\sum_{i=1}^{k}\theta_i{y}^i \leq gz$.
\end{proof}
%\cindy{TODO: Is this something that should be clear to the reader with what they know at this point? Does it need an explanation? Does it need an explanation at the end of this section or the next?}\arash{this is just a relabeling of the indices.} \cindy{Sorry.  I didn't understand that you meant the solution has that property. Right? The vector $x$ isn't defined in this paragraph and it's not in the next lemma either. If $x$ is the output of DomToIP, then we should say that.  ``Without loss of generality the solution $x$ returned by DomToIP has ...''}\arash{I see your point, please see if the change made it better} \cindy{Yes, thanks.  We can delete this conversation after you see this response.}

>>>>

Consider an IP instance $I=(A,b)$. Define sets $S(I)$ and $P(I)$ as in (\ref{S}) and (\ref{P}), respectively. Assume $S(I)\subseteq \{0,1\}^n$ and $P(I)\subseteq [0,1]^n$. For simplicity in the notation we assume an instance $I$ and denote $P(I),S(I),$ and $g(I)$ by $P$, $S$, and $g$ for this section and the next section. Also, for both sections let $x^*$ be the optimal solution to the LP relaxation and assume $t=|\spp(x^*)|$. Reorder variables as necessary to put all the nonzero variables into the first $t$ indices. So we can assume $x^*_i = 0$ for $i=t+1,\ldots,n$.

Before diving into the main results, we state the following simple observation that we will use several times in the proofs of this section and the next section.
\begin{observation}\label{dom01}
Any point $z\in [0,1]^n$ that is a convex combination of points in $\dom(S)$, is also a convex combination of points in $\dom(S) \cap [0,1]^n$
\end{observation}
\begin{proof}
This is because if there exists ${z}^i\in \dom(S)$ and $\theta_i\geq 0$ for $i\in [k]$ such that $\sum_{i=1}^{k} \theta_i = 1$ and $\sum_{i=1}^{k}\theta_i {z}^i \leq gz$, then 
we let ${y}^i$ be the coordiante-wise minimum of vector ${z}^i$ and $\{1\}^n$ for $i\in [k]$ and we have ${y}^i\in \dom(S)$ and $\sum_{i=1}^{k}\theta_i{y}^i \leq gz$.
\end{proof}
%\cindy{TODO: Is this something that should be clear to the reader with what they know at this point? Does it need an explanation? Does it need an explanation at the end of this section or the next?}\arash{this is just a relabeling of the indices.} \cindy{Sorry.  I didn't understand that you meant the solution has that property. Right? The vector $x$ isn't defined in this paragraph and it's not in the next lemma either. If $x$ is the output of DomToIP, then we should say that.  ``Without loss of generality the solution $x$ returned by DomToIP has ...''}\arash{I see your point, please see if the change made it better} \cindy{Yes, thanks.  We can delete this conversation after you see this response.}
